---
title: "Some prior predictive checks"
format:
  html:
    code-fold: true
output: pdf_document
---

```{r, message=FALSE}
library(tidyverse)
source("../../init.R")
```

Navigating to the US directory to load use the US data loading wrappers.

```{r}
find_and_set_directory("US/exploration")
source("load_and_preprocess.R")
find_and_set_directory("US/exploration")
source("compute_descriptives.R")
```

In particular, I want to focus on two tract states, Ohio and Massachusetts, as well as two ZIP states, New Jersey and Illinois.

## Ohio

```{r}
data_summary("OH", year = 2010)
```


I take one additional preprocessing steps on the predictors passed to the logit side (income, building period, and SVI), I also standardise the pediatrician rate.

```{r}
# standardise pediatrician rate
oh_merged <- oh_merged |>
  mutate(ped_per_100k = (ped_per_100k - mean(ped_per_100k)) / sd(ped_per_100k))
```


```{r}
# load STAN model
library(cmdstanr)

find_and_set_directory("models/prior_predictive_checks")
ppc_stan <- cmdstan_model("poisson_thinned_exclusion_ppc.stan")

# create function to prep data for given STAN model

stan_data <- function(state_final){
  list(
    N_obs = state_final |> filter(!BLL_geq_5_suppressed) |> count() |> pull(n),
    N_cens = state_final |> filter(BLL_geq_5_suppressed) |> count() |> pull(n),
    y_obs = state_final |> filter(!BLL_geq_5_suppressed) |> pull(BLL_geq_5),
    # obs
    median_income_obs = state_final |> filter(!BLL_geq_5_suppressed) |> pull(median_annual_incomeE),
    house_price_obs = state_final |> filter(!BLL_geq_5_suppressed) |> pull(house_price_medianE),
    poverty_obs = state_final |> filter(!BLL_geq_5_suppressed) |> pull(poor_fam_propE),
    black_prop_obs = state_final |> filter(!BLL_geq_5_suppressed) |> pull(black_ppl_propE),
    building_period_obs = state_final |> filter(!BLL_geq_5_suppressed) |> pull(bp_pre_1959E_prop),
    svi_obs = state_final |> filter(!BLL_geq_5_suppressed) |> pull(svi_socioeconomic_pctile),
    # censored
    median_income_cens = state_final |> filter(BLL_geq_5_suppressed) |> pull(median_annual_incomeE),
    house_price_cens = state_final |> filter(BLL_geq_5_suppressed) |> pull(house_price_medianE),
    poverty_cens = state_final |> filter(BLL_geq_5_suppressed) |> pull(poor_fam_propE),
    black_prop_cens = state_final |> filter(BLL_geq_5_suppressed) |> pull(black_ppl_propE),
    building_period_cens = state_final |> filter(BLL_geq_5_suppressed) |> pull(bp_pre_1959E_prop),
    svi_cens = state_final |> filter(BLL_geq_5_suppressed) |> pull(svi_socioeconomic_pctile),
    # pediatricians & kids
    z_obs = state_final |> filter(!BLL_geq_5_suppressed) |> pull(ped_per_100k),
    z_cens = state_final |> filter(BLL_geq_5_suppressed) |> pull(ped_per_100k),
    kids_obs = state_final |> filter(!BLL_geq_5_suppressed) |> pull(under_yo5_pplE),
    kids_cens = state_final |> filter(BLL_geq_5_suppressed) |> pull(under_yo5_pplE)
  )
}
```
Setting the priors as follows:

In the logit:
- intercept of N(0, 1.2) for "uniform" log-odds
- slopes of N(0,0.5) for all 4 predictors (income, building period, SVI, and pediatrician rate)
In the poisson:
- intercept of N(-1.58,2) based on the NHANES national average of 2.6%
- slopes of N(0,0.5) for all intercepts - for higher variances some observations go beyond the rate tolerance - except income and the building period, which I center -0.1 and 0.1 respectively.

```{r}
# plot N(0, 0.6)
plot(density(rnorm(500,0,0.4)))
```


```{r}
# sample from prior predictive distribution
oh_stan <- stan_data(oh_merged)

ppc_samples <- ppc_stan$sample(
  data = oh_stan,
  iter_sampling = 500,
  refresh = NULL,
  fixed_param = TRUE) # required for RNG?
```
```{r}
ppc_samples$summary(variables = c("y_thinned"))
```


```{r}
# add y_thinned from $summary to oh_merged
oh_merged_ppc <- oh_merged |> 
  bind_cols(ppc_samples$summary(variables = c("y_thinned")) |> 
              filter(str_detect(variable, "y_thinned")) |>
              select(c(mean, median, q5, q95)))
```

Plot the thinning rates

```{r}
library(bayesplot)

ppc_samples$draws(format = "draws_df") |> 
  # select columns that contain "pi"
  select(starts_with("pi")) |>
  # randomly select 100 columns of those (for speed)
  select(sample(1:nrow(oh_merged_ppc), 10)) |>
  mcmc_areas(prob = 0.9) +
  ggtitle("Prior thinning rates for Ohio")
```



```{r}
library(bayesplot)

# extract and plot draws
ppc_samples$draws(format = "draws_df", variables = c("y_thinned")) |> 
  # select columns that contain "y_thinned"
  select(starts_with("y_thinned")) |>
  # randomly select 100 columns of those (for speed)
  select(sample(1:length(oh_merged), 10)) |>
  mcmc_areas(prob = 0.8, alpha = 0.1) +
  ggtitle("Prior predictive checks for Ohio") +
  xlim(0,30)
```


## Massachusetts

```{r}
data_summary("MA", year = 2010)
```

```{r}
# standardise pediatrician rate
ma_merged <- ma_merged |>
  mutate(ped_per_100k = (ped_per_100k - mean(ped_per_100k)) / sd(ped_per_100k))
```

```{r}
# sample from prior predictive distribution
ma_stan <- stan_data(ma_merged)

ppc_samples <- ppc_stan$sample(
  data = ma_stan,
  iter_sampling = 500,
  refresh = NULL,
  fixed_param = TRUE)
```

```{r}
# add y_thinned from $summary to ma_merged
ma_merged_ppc <- ma_merged |> 
  bind_cols(ppc_samples$summary(variables = c("y_thinned")) |> 
              filter(str_detect(variable, "y_thinned")) |>
              select(c(mean, median, q5, q95)))
```

```{r}
# extract prior thinning rates
ppc_samples$draws(format = "draws_df") |> 
  # select columns that contain "pi"
  select(starts_with("pi")) |>
  # randomly select 100 columns of those (for speed)
  select(sample(1:nrow(ma_merged), 10)) |>
  mcmc_areas(prob = 0.9) +
  ggtitle("Prior thinning rates for Massachusetts")
```


```{r}
# extract and plot draws
ppc_samples$draws(format = "draws_df", variables = c("y_thinned")) |> 
  # select columns that contain "y_thinned"
  select(starts_with("y_thinned")) |>
  # randomly select 100 columns of those (for speed)
  select(sample(1:nrow(ma_merged_ppc), 10)) |>
  mcmc_areas(prob = 0.9, alpha = 0.1) +
  ggtitle("Prior predictive checks for Massachusetts") +
  xlim(0,20)
```

- 