---
title: "Descriptives of tract states"
format:
  pdf:
    toc: true
    code-fold: true
---

```{r, message=FALSE, warning=FALSE}
library(kableExtra)
library(tidyverse)
library(jsonlite)
source("../../init.R")
source("load_and_preprocess.R")
```

Getting metadata
```{r, error=TRUE}
# get metadata from ~US root
metadata  <- fromJSON("../metadata.json")
# extract required metadata (lists)
zip_states <- metadata$zip_states
tract_states <- metadata$tract_states
```

```{r, error=TRUE}
source("compute_descriptives.R")
```

# TRACT states

```{r, error=TRUE}
print(tract_states)
```

## Ohio

```{r, error=TRUE}
oh_summary <- data_summary("OH", year = NULL)
```

```{r, error=TRUE}
oh_summary[[1]] |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
oh_summary[[2]] 
```

```{r, error=TRUE}
oh_summary[[3]]
```

```{r, error=TRUE}
oh_summary[[4]]
```

```{r, error=TRUE}
oh_lead_model <- lead_count_model_summary("OH", year = 2010)
```

```{r, error=TRUE}
oh_lead_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
oh_lead_model$plot
```

```{r, error=TRUE}
oh_test_model <- test_count_model_summary("OH", year = 2012)
```

```{r, error=TRUE}
oh_test_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
oh_test_model$plot
```

## Pennsylvania

```{r, error=TRUE}
pa_summary <- data_summary("PA", year = NULL)
```

```{r, error=TRUE}
pa_summary[[1]] |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
pa_summary[[2]] 
```

```{r, error=TRUE}
pa_summary[[3]]
```

```{r, error=TRUE}
pa_summary[[4]]
```

```{r, error=TRUE}
pa_lead_model <- lead_count_model_summary("PA", year = 2007)
```

```{r, error=TRUE}
pa_lead_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
pa_lead_model$plot
```

```{r, error=TRUE}
pa_test_model <- test_count_model_summary("PA", year = 2007)
```

```{r, error=TRUE}
pa_test_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
pa_test_model$plot
```

## Colorado

```{r, error=TRUE}
co_summary <- data_summary("CO", year = NULL)
```

```{r, error=TRUE}
co_summary[[1]] |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
co_summary[[2]] 
```

```{r, error=TRUE}
co_summary[[3]]
```

```{r, error=TRUE}
co_summary[[4]]
```

```{r, error=TRUE}
co_lead_model <- lead_count_model_summary("CO", year = 2014)
```

```{r, error=TRUE}
co_lead_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
co_lead_model$plot
```

```{r, error=TRUE}
co_test_model <- test_count_model_summary("CO", year = 2014)
```

```{r, error=TRUE}
co_test_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
co_test_model$plot
```

## Maryland

```{r, error=TRUE}
md_summary <- data_summary("MD", year = NULL)
```

```{r, error=TRUE}
md_summary[[1]] |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
md_summary[[2]] 
```

```{r, error=TRUE}
md_summary[[3]]
```

```{r, error=TRUE}
md_summary[[4]]
```

```{r, error=TRUE}
md_lead_model <- lead_count_model_summary("MD", year = 2011)
```

```{r, error=TRUE}
md_lead_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
md_lead_model$plot
```

```{r, error=TRUE}
md_test_model <- test_count_model_summary("MD", year = 2011)
```

```{r, error=TRUE}
md_test_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
md_test_model$plot
```

## Massachusetts

```{r, error=TRUE}
ma_summary <- data_summary("MA", year = NULL)
```

```{r, error=TRUE}
ma_summary[[1]] |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
ma_summary[[2]] 
```

```{r, error=TRUE}
ma_summary[[3]]
```

```{r, error=TRUE}
ma_summary[[4]]
```

```{r, error=TRUE}
ma_lead_model <- lead_count_model_summary("MA", year = 2014)
```

```{r, error=TRUE}
ma_lead_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
ma_lead_model$plot
```

```{r, error=TRUE}
ma_test_model <- test_count_model_summary("MA", year = 2014)
```

```{r, error=TRUE}
ma_test_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
ma_test_model$plot
```

## Minnesota

```{r, error=TRUE}
mn_summary <- data_summary("MN", year = NULL)
```

```{r, error=TRUE}
mn_summary[[1]] |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
mn_summary[[2]] 
```

```{r, error=TRUE}
mn_summary[[3]]
```

```{r, error=TRUE}
mn_summary[[4]]
```

```{r, error=TRUE}
mn_lead_model <- lead_count_model_summary("MN", year = 2014)
```

```{r, error=TRUE}
mn_lead_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
mn_lead_model$plot
```

```{r, error=TRUE}
mn_test_model <- test_count_model_summary("MN", year = 2014)
```

```{r, error=TRUE}
mn_test_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
mn_test_model$plot
```

NYC is not available.

## North Carolina

```{r, error=TRUE}
nc_summary <- data_summary("NC", year = NULL)
```

```{r, error=TRUE}
nc_summary[[1]] |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
nc_summary[[2]] 
```

```{r, error=TRUE}
nc_summary[[3]]
```

```{r, error=TRUE}
nc_summary[[4]]
```

```{r, error=TRUE}
nc_lead_model <- lead_count_model_summary("NC", year = 2014)
```

```{r, error=TRUE}
nc_lead_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
nc_lead_model$plot
```

```{r, error=TRUE}
nc_test_model <- test_count_model_summary("NC", year = 2014)
```

```{r, error=TRUE}
nc_test_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
nc_test_model$plot
```

## Indiana

```{r, error=TRUE}
in_summary <- data_summary("IN", year = NULL)
```

```{r, error=TRUE}
in_summary[[1]] |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
in_summary[[2]] 
```

```{r, error=TRUE}
in_summary[[3]]
```

```{r, error=TRUE}
in_summary[[4]]
```

```{r, error=TRUE}
in_lead_model <- lead_count_model_summary("IN", year = 2013)
```

```{r, error=TRUE}
in_lead_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
in_lead_model$plot
```

```{r, error=TRUE}
in_test_model <- test_count_model_summary("IN", year = 2013)
```

```{r, error=TRUE}
in_test_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
in_test_model$plot
```

## Oregon

```{r, error=TRUE}
or_summary <- data_summary("OR", year = NULL)
```

```{r, error=TRUE}
or_summary[[1]] |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
or_summary[[2]] 
```

```{r, error=TRUE}
or_summary[[3]]
```

```{r, error=TRUE}
or_summary[[4]]
```

```{r, error=TRUE}
or_lead_model <- lead_count_model_summary("OR", year = 2010)
```

```{r, error=TRUE}
or_lead_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
or_lead_model$plot
```

```{r, error=TRUE}
or_test_model <- test_count_model_summary("OR", year = 2010)
```

```{r, error=TRUE}
or_test_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
or_test_model$plot
```

## New Hampshire

```{r, error=TRUE}
nh_summary <- data_summary("NH", year = NULL)
```

```{r, error=TRUE}
nh_summary[[1]] |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
nh_summary[[2]] 
```

```{r, error=TRUE}
nh_summary[[3]]
```

```{r, error=TRUE}
nh_summary[[4]]
```

```{r, error=TRUE}
nh_lead_model <- lead_count_model_summary("NH", year = 2010)
```

```{r, error=TRUE}
nh_lead_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
nh_lead_model$plot
```

```{r, error=TRUE}
nh_test_model <- test_count_model_summary("NH", year = 2010)
```

```{r, error=TRUE}
nh_test_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
nh_test_model$plot
```

## Wisconsin

```{r, error=TRUE}
wi_summary <- data_summary("WI", year = NULL)
```

```{r, error=TRUE}
wi_summary[[1]] |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
wi_summary[[2]] 
```

```{r, error=TRUE}
wi_summary[[3]]
```

```{r, error=TRUE}
wi_summary[[4]]
```

```{r, error=TRUE}
wi_lead_model <- lead_count_model_summary("WI", year = 2007)
```

```{r, error=TRUE}
wi_lead_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
wi_lead_model$plot
```

```{r, error=TRUE}
wi_test_model <- test_count_model_summary("WI", year = 2007)
```

```{r, error=TRUE}
wi_test_model$table |> knitr::kable() |> kable_styling(latex_options = "scale_down")
```

```{r, error=TRUE}
wi_test_model$plot
```

# Model summaries
## Lead counts
For all (available) tract states, we plot the coefficients of the lead count model.

```{r, error=TRUE}
# List of lead model names
lead_model_names <- c("oh_lead_model", "pa_lead_model", "co_lead_model", "md_lead_model", "ma_lead_model", "mn_lead_model", "nc_lead_model", "in_lead_model", "or_lead_model", "nh_lead_model", "wi_lead_model")

# Safely retrieve the fitted lead models
lead_models <- mget(lead_model_names, envir = .GlobalEnv, ifnotfound = list(NULL))

# Filter out any NULL values in models and their names
lead_model_names <- lead_model_names[!sapply(lead_models, is.null)]
lead_models <- lead_models[!sapply(lead_models, is.null)]

# Function to safely extract and filter data
safe_extract <- possibly(function(model, variable) {
  model$plot |> 
    pluck(1) |> 
    filter(parameter == variable)
}, otherwise = data.frame())

# Extract and combine the density data
get_variable_densities <- function(variable){
    map2_dfr(lead_models, lead_model_names, ~ {
    data <- safe_extract(.x, variable)
    if (nrow(data) > 0) {
      data$state <- .y
    }
    data
    })
}
```

### Median annual income
```{r, fig.height = 12, fig.width = 8}
# Get the density data for median annual income
density_data_income <- get_variable_densities("median_annual_incomeE")

# Create the plot
ggplot(density_data_income, aes(x = x, y = density, fill = state)) + 
  geom_area(alpha = 0.6, position = "identity") + 
  facet_wrap(~ state, scales = "free_y", ncol = 1) + 
  theme_minimal() + 
  labs(title = "Density of Median Annual Income by State",
       x = "Income",
       y = "Density")
```

### House price median

```{r, fig.height = 12, fig.width = 8}
# Get the density data for house price median
density_data_house_price <- get_variable_densities("house_price_medianE")

# Create the plot
ggplot(density_data_house_price, aes(x = x, y = density, fill = state)) + 
  geom_area(alpha = 0.6, position = "identity") + 
  facet_wrap(~ state, scales = "free_y", ncol = 1) + 
  theme_minimal() + 
  labs(title = "Density of House Price Median by State",
       x = "House Price",
       y = "Density")
```

### Building period

```{r, fig.height = 12, fig.width = 8}
# Get the density data for building period
density_data_building_period <- get_variable_densities("bp_pre_1959E_prop")

# Create the plot
ggplot(density_data_building_period, aes(x = x, y = density, fill = state)) + 
  geom_area(alpha = 0.6, position = "identity") + 
  facet_wrap(~ state, scales = "free_y", ncol = 1) + 
  theme_minimal() + 
  labs(title = "Density of Building Period Pre-1959 by State",
       x = "Building Period Pre-1959",
       y = "Density")
```


## Test counts

For all (available) tract states, we plot the coefficients of the test count model.

```{r}
test_model_names <- c("oh_test_model", "pa_test_model", "co_test_model", "md_test_model", "ma_test_model", "mn_test_model", "nc_test_model", "in_test_model", "or_test_model", "nh_test_model", "wi_test_model")

test_models <- mget(test_model_names, envir = .GlobalEnv, ifnotfound = list(NULL))

test_model_names <- test_model_names[!sapply(test_models, is.null)]
test_models <- test_models[!sapply(test_models, is.null)]

get_variable_densities_test <- function(variable){
    map2_dfr(test_models, test_model_names, ~ {
    data <- safe_extract(.x, variable) # can use same extraction function
    if (nrow(data) > 0) {
      data$state <- .y
    }
    data
    })
}
```

### Pediatricians

```{r, fig.height = 12, fig.width = 8}
density_data_pediatricians_test <- get_variable_densities_test("ped_per_100k")

ggplot(density_data_pediatricians_test, aes(x = x, y = density, fill = state)) + 
  geom_area(alpha = 0.6, position = "identity") + 
  facet_wrap(~ state, scales = "free_y", ncol = 1) + 
  theme_minimal() + 
  labs(title = "Density of Pediatricians per 100k by State",
       x = "Pediatricians per 100k",
       y = "Density")
```

