This notebook analyses some of the suppression issues present in the cleaned data files.

```{r}
library(tidyverse)
library(ggplot2)
library(readxl)
```

# Motivating questions:

1.  Find out if any zips / tracts suppress counts of tests in addition to counts of elevated BLLs. I think the answer is yes, but this may potentially vary by State.
2.  Create summaries / counts / histograms that explain the pattern and extent of suppressed counts, including:

<!-- -->

a.  How many zips / tracts are suppressed in each state?
b.  If we were to aggregate tracts to zips, how often would we be aggregating 1 suppressed, 2 suppressed, etc. units. I'm not sure the best way of displaying this. Maybe a histogram?

<!-- -->

3.  Make a spreadsheet with the state-level conventions for cell suppression: what is suppressed, and what is the cutoff? Any other relevant information?

The answer to 1 & 3 is contained in the excel sheet in which we collect all information regarding raw BLL data files: "description-of-raw-BLL_data.xlsx". **This notebook will focus on 2.**

## Read in Excel sheet

```{r}
# read in "description-of-raw-BLL_data.xlsx"
desc <- read_excel("../description-of-raw-BLL-data.xlsx", sheet = 1) |> 
  select(State, "Reporting Level", Suppression) 

desc
```

## ZIP level

From the table we already know that among the zip-level states, some do not suppress any data:

```{r}
desc |>
  rename(level = "Reporting Level") |> 
  filter((level == "zip") & (Suppression == "FALSE"))
```

```{r}
# load and merge all ZIP states
# setwd to source files
setwd("../scripts/source files/")
source("../00_merging_functions.R")
load_states(zip_states)
```

Merging:

```{r}
merged <- merge_loaded_data(str_to_lower(zip_states))
# get rid of individual zip state files
rm(list = str_to_lower(zip_states))
```

The combined zip data has around 290k observations.

How many zips are suppressed in each state?

```{r}
merged <- merged |> 
  # check suppression using whether first character is "<"
  mutate(BLL_geq_5_sup = str_detect(BLL_geq_5, "<"),
         BLL_geq_10_sup = str_detect(BLL_geq_10, "<"),
         tested_sup = str_detect(tested, "<"))
```

```{r}
suppr_summ <- merged |> 
  group_by(state) |> 
  summarize(n_zips = n(),
            n_BLL_5_sup = sum(BLL_geq_5_sup, na.rm = TRUE),
            n_BLL_10_sup = sum(BLL_geq_10_sup, na.rm = TRUE),
            n_tested_sup = sum(tested_sup, na.rm = TRUE)) |> 
  mutate(perc_BLL_5_sup = n_BLL_5_sup / n_zips,
         perc_BLL_10_sup = n_BLL_10_sup / n_zips,
         perc_tested_sup = n_tested_sup / n_zips)

```

There can be suppression both in "tested" as well as in the BLL counts. If the number of tested is suppressed, the count of EBLLs is also suppressed or 0. For 13 states, there is suppression also at the testing level:

```{r}
# display results sorted by tested_sup
suppr_summ |> 
  arrange(desc(perc_tested_sup)) |> 
  relocate(perc_tested_sup, .after = state)
```

```{r}
# plot a scatter of the suppressed tested and BLL_geq_5 ratios
suppr_summ |>
  filter((perc_BLL_5_sup != 0) & (perc_tested_sup != 0)) |> 
  ggplot(aes(x = perc_tested_sup, y = perc_BLL_5_sup)) +
  geom_point() +
  labs(x = "% of ZIPs with suppressed tested counts",
       y = "% of ZIPs with suppressed BLL_geq_5 counts",
       title = "Scatter of ZIP suppression rates") +
  geom_text(aes(label = state), nudge_x = 0.03, nudge_y = - 0.03) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_bw()
```

## Tract level

```{r}
setwd("../scripts/source files/")
load_states(tract_states)
```

The combined tract level data has around 300k observations:

```{r}
merged_tract <- merge_loaded_data(str_to_lower(tract_states), level = "tract")
# remove individual states
rm(list = str_to_lower(tract_states))
dim(merged_tract)
```

Let's look at the same suppression statistics as for zips.

First, note that again, there are some states that report at tracts and do not suppress at all:

```{r}
desc |>
  rename(level = "Reporting Level") |> 
  filter((str_detect(str_to_lower(level), "tract")) & (Suppression == "FALSE"))
```

For the others, suppression rates vary:

```{r}
merged_tract <- merged_tract |> 
  # check suppression using whether first character is "<"
  mutate(BLL_geq_5_sup = str_detect(BLL_geq_5, "<"),
         BLL_geq_10_sup = str_detect(BLL_geq_10, "<"),
         tested_sup = str_detect(tested, "<"))

```

```{r}
# get summary stats
suppr_summ_tract <- merged_tract |> 
  group_by(state) |> 
  summarize(n_tracts = n(),
            n_BLL_5_sup = sum(BLL_geq_5_sup, na.rm = TRUE),
            n_BLL_10_sup = sum(BLL_geq_10_sup, na.rm = TRUE),
            n_tested_sup = sum(tested_sup, na.rm = TRUE)) |> 
  mutate(perc_BLL_5_sup = n_BLL_5_sup / n_tracts,
         perc_BLL_10_sup = n_BLL_10_sup / n_tracts,
         perc_tested_sup = n_tested_sup / n_tracts)
```

```{r}
# display results sorted by tested_sup
suppr_summ_tract |> 
  arrange(desc(perc_tested_sup)) |> 
  relocate(perc_tested_sup, .after = state)
```

```{r}
# plot a scatter of the suppressed tested and BLL_geq_5 ratios
suppr_summ_tract |>
  filter((perc_BLL_5_sup != 0) & (perc_tested_sup != 0)) |> 
  ggplot(aes(x = perc_tested_sup, y = perc_BLL_5_sup)) +
  geom_point() +
  labs(x = "% of tracts with suppressed tested counts",
       y = "% of tracts with suppressed BLL_geq_5 counts",
       title = "Scatter of ZIP suppression rates") +
  geom_text(aes(label = state), nudge_x = 0.03, nudge_y = - 0.03) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_bw()
```

Oregon, for example, has **all** of its tracts with suppressed BLL counts.

### Aggregating tracts to ZIPs

We cannot use the wrapper from the *00_tract-zip_crosswalk_function.R* since this is meant to also wrap the aggregation of the tested & BLL variables, which requires a numeric type. At this point, we want to keep suppression information in its raw form, so we proceed "by hand":

```{r}
setwd("../raw_files")
tract_path <- 'TRACT_ZIP_032010.xlsx'
drop_get_from_root(tract_path)

tracttozip <- read_excel(tract_path) %>% 
  rename(tract=TRACT)
```

Merge the crosswalking file with

```{r}
merged_tract <- merged_tract |> 
  left_join(tracttozip, by = "tract")
```

First, let's check how good coverage of zips is (in terms of the count):

```{r}
merged_tract |> 
  group_by(state) |> 
  # count NAs in ZIP
  summarize(n_tracts = n(),
            n_NA_zip = sum(is.na(ZIP)),
            perc_tracts_lost = n_NA_zip / n_tracts) |> 
  arrange(desc(perc_tracts_lost))
```

Plotting:

```{r}
merged_tract |> 
  group_by(state) |> 
  # count NAs in ZIP
  summarize(n_tracts = n(),
            n_NA_zip = sum(is.na(ZIP)),
            perc_tracts_lost = n_NA_zip / n_tracts) |> 
  arrange(desc(perc_tracts_lost)) |> 
  ggplot(aes(x = state, y = perc_tracts_lost)) +
  geom_bar(stat = "identity") +
  labs(title = "Percentage of tracts not matched to any ZIP by state",
       y = "%")
```

In NYC, this is due to a data issue in how tracts are reported (and how its currently cleaned.)

However, the amount of tracts lost is not insignificant for the other states either. With the current cross-walking file, overall, one loses roughly 18% of the data:

```{r}
merged_tract |> 
  summarize(n_tracts = n(),
            n_NA_zip = sum(is.na(ZIP)),
            perc_tracts_lost = n_NA_zip / n_tracts) |> 
  pull(perc_tracts_lost)
```

Now, let's turn to the second part of the question. If one were to proceed with aggregation to ZIPs, how many ZIPs would contain which nr. of suppressed tracts, and what would their cumulative RES_RATIO be?

```{r}
agg_results <- merged_tract |> 
  mutate(any_sup = BLL_geq_5_sup | BLL_geq_10_sup | tested_sup) |>
  group_by(ZIP) |>
  summarize(n_tracts = n(),
            n_BLL_5_sup = sum(BLL_geq_5_sup, na.rm = TRUE),
            n_BLL_10_sup = sum(BLL_geq_10_sup, na.rm = TRUE),
            n_tested_sup = sum(tested_sup, na.rm = TRUE),
            n_any_sup = sum(any_sup, na.rm = TRUE),
            perc_any_sup = n_any_sup / n_tracts,
            # get sum of RES RATIO for all tracts
            cum_RES_RATIO = sum(RES_RATIO, na.rm = TRUE),
            # get sum of RES RATIO for suppressed tracts
            cum_RES_RATIO_sup = sum(RES_RATIO[any_sup], na.rm = TRUE)) |> 
  arrange(desc(perc_any_sup)) |> 
  relocate(perc_any_sup, .after = ZIP)

agg_results
```

Plotting the results

```{r}
agg_results |> 
  ggplot(aes(x = perc_any_sup)) +
  geom_histogram(bins = 20) +
  labs(title = "Count of ZIPs by percentage of suppressed tracts",
       x = "% of tracts with suppression in ANY variable",
       y = "Count")
```

"Good news" is that there is a large mass on 0: These are ZIPS for which NONE of the tracts in it have any sort of suppression.
