This notebook explores reducing the shapefile size to speed up interactive mapping of England MSOAs.

```{r}
library(tidyverse)
library(rmapshaper)
library(sf)
```

First, load the shapefile using sf

```{r}
msoa <- st_read("../raw_data/shapefiles/msoa/england_msoa_2011.shp")
```

Let's take Birmingham as an example, since it has a lot of small, intricate MSOAs. Let's plot it to see what it looks like.

```{r}
birmingham <- msoa %>% filter(grepl("Birmingham", name))

birmingham |> 
    ggplot() +
    geom_sf()
```

Now let's make use of the rmapshaper package to simplify the shapefile. We'll use the Visvalingam-Whyatt algorithm, and keep 5% of the original points, which is the default.

```{r}
birmingham_simple <- ms_simplify(birmingham, keep = 0.05)
```

Let's again plot the area of Birmingham to see what it looks like.

```{r}
birmingham_simple |> 
    ggplot() +
    geom_sf()
```

For plotting purposes, the difference is barely noticeable, even though we kept only 5% of observations!

**NOTE:** Running this algorithm on the entire shapefile can lead the underlying mapshaper engine to run out of memory (see [here](https://github.com/ateucher/rmapshaper/issues/83#issuecomment-492775206)). One can alternatively run this on system using the `sys = TRUE` keyword, but it requires install node.

Instead, we can do it in chunks:

```{r}
library(purrr)

# assign chunks of ~ 200 rows
msoa_simple <- msoa |> 
    mutate(chunk = ceiling(row_number() / 200)) |> 
    group_by(chunk) |> 
    nest() |> 
    mutate(data = purrr::map(data, ms_simplify)) |> 
    unnest()
```

This way we avoided any Memory issues.

```{r}
# saving the simple shapefile
st_write(msoa_simple, "../raw_data/shapefiles/msoa/england_msoa_2011_simple.shp")
```
